Joint estimation of ancestral states and phylogenetic model parameters is inconsistent

### 2017-11-15 appendix.tex
fix up arguments for restricted case; disjoint ideas for general case
a few ideas trying to work through general case
Simplification and prepping for a bio future.
get all content into main doc

\section{Appendix}

\subsection{Likelihood compuations}

@@

>>>>

\section*{Appendix}

\subsection*{Hadamard transform}

Define: $\mathcal{T}$, $X$, $A$, $n$, $P$ and $\theta$.
Let
$$
\theta(e) = 1-2p(e).
$$
Then
$$
p_A = \frac{1}{2^{n-1}} \ \sum_{Y \subset X : |Y| \equiv 0 (\mathrm{mod} \ 2)} \ \left[(-1)^{|Y \cap A|} \ \prod_{e \in P(\mathcal{T}, Y)} \ \theta(e) \right]
$$

\subsection*{Site split formulation}

Since we have a finite character alphabet, for a given column $i$ there are a finite number of possible assignments of characters to tips $\alignmentColumn_i$ or internal nodes $\ancestralStateColumn_i$; this results in a simplification of likelihood calculation (we follow Section 8.6 of \cite{Semple2003-em}).
Take the tip labels of $\tau$ to be $\{1,\ldots,\nSiteRows\}$.
For likelihood calculation under the binary symmetric model, we describe a given $\alignmentColumn_i$ as a subset of indices $\siteSplit\subseteq\siteSplitSet:=\{1,\ldots,\nSiteRows-1\}$ with equivalent characters, commonly called a ``site split.''
Indeed, we start by including those tip labels in $\siteSplit$ that are assigned a 1 in the $\alignmentColumn_i$.
Then, because the probability of observing a particular collection of binary characters is equivalent to the probability of its complement under the binary symmetric model, we take the complement if needed to exclude label $\nSiteRows$ from $\siteSplit$.

For topology $\tau$, we define an ordered set of internal node labels $\{1,\ldots,\nAncestralStateRows\}$ for $\ancestralStateColumn_i$ and similarly use a subset of characters $\ancestralSplit\subseteq\ancestralSplitSet:=\{1,\ldots,\nAncestralStateRows\}$ to describe a realization of $\ancestralStateColumn_i$.
In this case the entire set of internal nodes must be enumerated: the probability of observing an ancestral state split, conditional on a site split, is not invariant to taking its complement.

We enumerate the site splits $\siteSplit_j$ of which there are $\nSiteSplits=|\mathcal{P}(\siteSplitSet)|$ in total where $\mathcal{P}$ denotes the power set.
Similarly we enumerate ancestral splits $\ancestralSplit_k$ of which there are $\nAncestralSplits=|\mathcal{P}(\ancestralSplitSet)|$ in total.

This split formulation defines mappings from patterns to splits as
 $$
 \patternToSplit:\alphabet^\nSiteRows\rightarrow\mathcal{P}(\siteSplitSet).
 $$
Given a site pattern--valued random variable $\alignmentColumnRV$, define the random variable
$$
\siteSplitRV := \patternToSplit(\alignmentColumnRV)
$$
that takes corresponding realizations $\siteSplit_j$ for some $j$.

For the $i$th factor of \eqref{eq:full_likelihood},
$$
P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) = P(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) \cdot P(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t).
$$
As a consequence of assuming a binary symmetric model, taking complements yields
\begin{align*}
    2\cdot P(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) &= P(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \\
                                                                 &= P(\siteSplitRV=\siteSplit_j \mid \tau, t)
\end{align*}
for some $j$.
For the other term, we define
$$
\ancestralToSplit:\alphabet^\nAncestralStateRows\times\alphabet^\nSiteRows\rightarrow\mathcal{P}(\ancestralSplitSet)
$$
as a mapping of ancestral states and tip states to ancestral state splits.
It is defined by whether the tip states have their complements taken or not: if a set of tip labels $\alignmentColumn$ is in $\siteSplitSet$, $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is $\ancestralStateColumn$; otherwise, if $\alignmentColumn$ is not in $\siteSplitSet$, then the complement of $\alignmentColumn$ necessarily is in $\siteSplitSet$, and $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is the complement of $\ancestralStateColumn$.
We define
$$
\ancestralSplitRV := \ancestralToSplit(\alignmentColumnRV, \ancestralStateColumnRV)
$$
for a tip state--valued random variable $\alignmentColumnRV$ and an ancestral state--valued random variable $\ancestralStateColumnRV$, so that
%By considering the case requiring complementing and that not requiring complementing separately,
%das: did I write the above? I don't remember the significance of it...
\begin{align*}
    P(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t) &= P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t).
\end{align*}
Given $(\tau, t)$, there exists an ordered list of sets $\fullAncestralSplitPartitions(\tau, t)=(\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t))$ such that any element $\xi_j$ of the $j$th component $\ancestralSplitPartition_j(\tau, t)$ satisfies
\begin{align*}
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t) &= P(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t).
\end{align*}
In other words, for the $j$th site split, $\ancestralSplitPartition_j(\tau, t)\subset\mathcal{P}(\ancestralSplitSet)$ is the set of most likely ancestral splits for that particular site split, topology and set of branch lengths, and $\xi_j$ is one of possibly many equiprobable ancestral state splits in $\ancestralSplitPartition_j(\tau, t)$.
For each $\alignmentColumn_i$, $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective, and from this have
\begin{align*}
\max_{\ancestralStateColumn_i} P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) &= P(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t).
\end{align*}

Let $\xi_j$ be such a choice for each $1 \leq j \leq q$ in the following.
Then, the likelihood in \eqref{eq:profile_likelihood} written as a product over site patterns as opposed to sites is
\begin{align}
L_\nCols'(\tau, t; \fullAlignment) &= \max_{\fullAncestralStates} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates) \nonumber \\
                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) \nonumber \\
                             &\propto \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ P(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \nonumber \\
                             &= \prod_{i=1}^{\nCols} \ P(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot \max_{\ancestralStateColumn_i} P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \nonumber \\
                             &= \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t)\cdot P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
\end{align}
where $\nCols_j(\fullAlignment)$ is the number of columns in $\fullAlignment$ where the site split $\siteSplit_j$ or its complement appears.

\paragraph{Example}
\begin{figure}
    \centering
    \includegraphics[width=.45\textwidth]{unrooted_farris_tree}
    \caption{Four taxon ``Farris zone'' tree.}
\label{fig:four-taxa-tree}
\end{figure}

Consider the fixed, binary four-tip tree $\tau$ in Fig.~\ref{fig:four-taxa-tree}---this is commonly known as the ``Farris zone'' topology.
The set of all possible character assignments is
\begin{align*}
\mathcal{P}(\{1,2,3,4\}) &= \{\emptyset, \{1,2,3,4\}, \{1\}, \{2,3,4\}, \{2\}, \{1,3,4\}, \{3\}, \{1,2,4\}, \\
                         &\qquad \{1,2\}, \{3,4\}, \{1,3\}, \{2,4\}, \{2,3\}, \{1,4\}, \{1,2,3\}, \{1,4\}\}.
\end{align*}
where each set indicates the tips assigned the character $1$.
For example, $\emptyset$ is the labeling $0000$ and $\{1,3,4\}$ is the labeling $1011$.
Symmetry allows us to group adjacent pairs in $\mathcal{P}(\{1,2,3,4\})$ into equiprobable splits, letting $\siteSplitSet=\{1,2,3\}$.
The unique site splits, collapsing complements, are
\begin{align*}
    \mathcal{P}(\siteSplitSet) &= \{\emptyset, \{1\}, \{2\}, \{3\}, \{1,2\}, \{1,3\}, \{2,3\}, \{1,2,3\}\} \\
& := \{\siteSplit_1, \ldots, \siteSplit_8\}.
\end{align*}
Since we identify character complements, we do not consider the additional splits
\begin{equation*}
\begin{split}
& \mathcal{P}(\{1,2,3,4\}) \setminus \mathcal{P}(\siteSplitSet) = \\
&\qquad \{\{1,2,3,4\}, \{2,3,4\}, \{1,3,4\}, \{1,2,4\}, \{3,4\}, \{2,4\}, \{1,4\}, \{4\}\},
\end{split}
\end{equation*}
the symmetry of the binary character model allowing us to focus only on the elements of $\mathcal{P}(\siteSplitSet)$.
This tree has two internal nodes with $\ancestralSplitSet=\{1,2\}$ and unique ancestral state splits
$$
\mathcal{P}(\ancestralSplitSet) = \{\emptyset, \{1\}, \{2\}, \{1,2\}\}.
$$
Internal node $\{1\}$ is the node connected to leaves $\{1\}$ and $\{3\}$ and internal node $\{2\}$ connected to leaves $\{2\}$ and $\{4\}$.
The mapping from characters to splits in this case will depend on the characters at the tips and the ancestral states.
For example, we take both $\patternToSplit(0000)=\emptyset$ and $\patternToSplit(1111)=\emptyset$.
Similarly, we have $\ancestralToSplit(0000, 00) = \emptyset$ and $\ancestralToSplit(1111, 11)=\emptyset$, needing to take the complement of all the characters present on the tree to identify splits.
We cannot identify complements for ancestral states in the same way as tip states since, for $\siteSplit\in\mathcal{P}(\siteSplitSet)$,
$$
P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\siteSplit, \tau, t)\neq P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\siteSplit, \tau, t)
$$
in general.

For each site split $\siteSplit\in\mathcal{P}(\siteSplitSet)$, we maximize the likelihood over all $\ancestralSplit\in\mathcal{P}(\ancestralSplitSet)$.
A maximum occurs at one of possibly several ancestral splits in $\mathcal{P}(\ancestralSplitSet)$, defined via $\ancestralSplitPartition_j(\tau, t)$ for the $j$th site split.
As a simple example, say all branch lengths correspond to a probability $p$ ($< 1/2$) of changing character along that branch.
The probabilities of observing ancestral splits for $\siteSplit_1=\emptyset$ are
$$
P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\emptyset, \tau, t) =
(1-p)^5,
$$
$$
P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\emptyset, \tau, t) =
P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\emptyset, \tau, t) =
p^3(1-p)^2,
$$
$$
P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\emptyset, \tau, t) =
p^4(1-p).
$$
The set of most likely ancestral states contains a single element, here $\ancestralSplitPartition_1(\tau, p)=\{\emptyset\}$.
Then, taking $\xi_1\in\ancestralSplitPartition_1(\tau, p)$ we have
$$
P(\ancestralSplitRV=\xi_1 \mid \siteSplitRV=\emptyset, \tau, t) =
P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\emptyset, \tau, t) =
(1-p)^5.
$$
For $\siteSplit_5=\{1,2\}$ we have
$$
P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1,2\}, \tau, t) =
P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
p^2(1-p)^3,
$$
$$
P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
p^3(1-p)^2.
$$
Here, the set of most likely ancestral states is $\ancestralSplitPartition_5(\tau, p)=\{\emptyset,\{1,2\}\}$, and, for $\xi_5\in\ancestralSplitPartition_5(\tau, p)$,
$$
P(\ancestralSplitRV=\xi_5 \mid \siteSplitRV=\{1,2\}, \tau, t) =
p^2(1-p)^3.
$$

\subsection*{Likelihood computations}

Table~\ref{tab:sitepatprob} contains calculations of site pattern frequencies under these two topologies.
Table~\ref{tab:likelihoods} contains calculations of likelihood values for fixed site patterns and topologies that have been maximized over ancestral state patterns.
Tables~\ref{tab:farris_likelihoods} and~\ref{tab:fels_likelihoods} contain the values of the likelihood for all possible ancestral states.

\begin{table}
\centering
\begin{tabular}{|l|l|l|}
    \hline
$\siteSplit_j$   &$P(\siteSplitRV=\siteSplit_j \mid \tau_1,t)$&$P(\siteSplitRV=\siteSplit_j \mid \tau_2,t)$\\
    \hline
    $\emptyset$    &$1+x'^2+y'^2+4x'y'w'+x'^2y'^2$&$1+2x'y'+2x'y'w'+x'^2w'+y'^2w'+x'^2y'^2$\\
    $\{1\}$        &$1-x'^2+y'^2-x'^2y'^2$&$1-x'^2w'+y'^2w'-x'^2y'^2$\\
    $\{2\}$        &$1+x'^2-y'^2-x'^2y'^2$&$1+x'^2w'-y'^2w'-x'^2y'^2$\\
    $\{3\}$        &$1-x'^2+y'^2-x'^2y'^2$&$1-x'^2w'+y'^2w'-x'^2y'^2$\\
    $\{1,2\}$      &$1-x'^2-y'^2+x'^2y'^2$&$1+2x'y'-2x'y'w'-x'^2w'-y'^2w'+x'^2y'^2$\\
    $\{1,3\}$      &$1+x'^2+y'^2-4x'y'w'+x'^2y'^2$&$1-2x'y'-2x'y'w'+x'^2w'+y'^2w'+x'^2y'^2$\\
    $\{2,3\}$      &$1-x'^2-y'^2+x'^2y'^2$&$1-2x'y'+2x'y'w'-x'^2w'-y'^2w'+x'^2y'^2$\\
    $\{1,2,3\}$    &$1+x'^2-y'^2-x'^2y'^2$&$1+x'^2w'-y'^2w'-x'^2y'^2$\\
    \hline
\end{tabular}
\caption{Site pattern probabilities.
All values are multiplied by $1/8$.}
\label{tab:sitepatprob}
\end{table}

\begin{table}
\centering
\begin{tabular}{|l|ll|}
    \multicolumn{3}{c}{$\tau=\tau_1$}\\
    \hline
    $\siteSplit_j$    & $\ancestralSplitPartition_j(\tau, t)$ & $P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j,\tau,t)$\\
    \hline
    $\emptyset$&
    $\emptyset$&
    $(1+x')^2   (1+w')(1+y')^2$\\
     $\{1\}$    &
    $\emptyset$&
    $(1+x')(1-x')(1+w')(1+y')^2$\\
     $\{2\}$    &
    $\emptyset$&
    $(1+x')^2   (1+w')(1+y')(1-y')$\\
     $\{3\}$    &
    $\emptyset$&
    $(1+x')(1-x')(1+w')(1+y')^2$\\
    $\{1,2\}$  &
    $\{\emptyset,\{1,2\}\}$&
    $(1+x')(1-x')(1+w')(1+y')(1-y')$\\
    $\{1,3\}$  &
    $\left\{\begin{array}{l}
                    \emptyset\\
                    \{1\}\\
                    \{1,2\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1-x')^2   (1+w')(1+y')^2\\
                    (1+x')^2   (1-w')(1+y')^2\\
                    (1+x')^2   (1+w')(1-y')^2
                \end{array}$\\
    $\{2,3\}$  &
                $\{\emptyset,\{1,2\}\}$&
                $(1+x')(1-x')(1+w')(1+y')(1-y')$\\
    $\{1,2,3\}$&
                $\{1,2\}$&
                $(1+x')^2   (1+w')(1+y')(1-y')$\\
    \hline
    \multicolumn{3}{c}{$\tau=\tau_2$}\\
    \hline
    $\siteSplit_j$    & $\ancestralSplitPartition_j(\tau, t)$ & $P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j,\tau,t)$\\
    \hline
    $\emptyset$       &$\emptyset$&$(1+x')^2   (1+w')(1+y')^2$\\
    $\{1\}$          &
    $\left\{\begin{array}{l}
                    \emptyset\\
                    \{1\}
                \end{array}\right.$&
    $\begin{array}{l}
                        (1+x')(1-x')(1+w')(1+y')^2\\
                        (1+x')^2   (1-w')(1+y')(1-y')
                    \end{array}$\\
      $\{2\}$          &
    $\left\{\begin{array}{l}
                    \emptyset\\
                    \{1\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x')^2   (1+w')(1+y')(1-y')\\
                    (1+x')(1-x')(1-w')(1+y')^2
                    \end{array}$\\
      $\{3\}$          &
    $\left\{\begin{array}{l}
                    \emptyset\\
                    \{2\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x')(1-x')(1+w')(1+y')^2\\
                    (1+x')^2   (1-w')(1+y')(1-y')
                    \end{array}$\\
     $\{1,2\}$         &
    $\left\{\begin{array}{l}
                    \{\emptyset,\{1,2\}\}\\
                    \{1\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x')(1-x')(1+w')(1+y')(1-y')\\
                    (1+x')^2   (1-w')(1+y')^2
                    \end{array}$\\
     $\{1,3\}$         &
    $\left\{\begin{array}{l}
                    \emptyset\\
                    \{\{1\},\{2\}\}\\
                    \{1,2\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1-x')^2   (1+w')(1+y')^2\\
                    (1+x')(1-x')(1-w')(1+y')(1-y')\\
                    (1+x')^2   (1+w')(1-y')^2
                    \end{array}$\\
      $\{2,3\}$        &
    $\left\{\begin{array}{l}
                    \{\emptyset,\{1,2\}\}\\
                    \{1\}\\
                    \{2\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x')(1-x')(1+w')(1+y')(1-y')\\
                    (1-x')^2   (1-w')(1+y')^2\\
                    (1+x')^2   (1-w')(1-y')^2
                    \end{array}$\\
     $\{1,2,3\}$       &
    $\left\{\begin{array}{l}
                    \{2\} \\
                    \{1,2\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x')(1-x')(1-w')(1+y')^2 \\
                    (1+x')^2   (1+w')(1+y')(1-y')
                    \end{array}$\\
    \hline
\end{tabular}
\caption{Likelihood calculations for all site patterns and ancestral state partitions.
All values multiplied by $1/32$.
Likelihoods with multiple entries have maxima determined by unknown branch length parameters.
See Appendix for full calculations.}
\label{tab:likelihoods}
\end{table}

@@

\subsection*{Form of the likelihood}

Using the Hadamard transform approach, we calculate the generating probabilities on the Farris tree topology.
For site split $\emptyset$,
\begin{align*}
    P(\siteSplitRV=\emptyset\mid \tau_1, \tilde{t}) & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  x_1y_1w + x_1y_2w + y_1x_2w + x_2y_2w + x_1y_1x_2y_2) \\
                                              & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1y_1 + x_1y_2 + y_1x_2 + x_2y_2] + x_1y_1x_2y_2) \\
                                              & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1 + x_2][y_1 + y_2] + x_1y_1x_2y_2).
\end{align*}
and this probability is unchanged when $x_1$ is exchanged with $x_2$ and $y_1$ is exchanged with $y_2$.
All other generating probabilities will differ only in the signs of each term.
For example, for site split $\{1\}$ we have
\begin{align*}
    P(\siteSplitRV=\{1\}\mid \tau_1, \tilde{t}) & = \frac{1}{8} (1 - x_1x_2 +  y_1y_2 +  w[-x_1 + x_2][y_1 + y_2] - x_1y_1x_2y_2)
\end{align*}
and for site split $\{3\}$ we have
\begin{align*}
    P(\siteSplitRV=\{3\}\mid \tau_1, \tilde{t}) & = \frac{1}{8} (1 - x_1x_2 +  y_1y_2 +  w[x_1 - x_2][y_1 + y_2] - x_1y_1x_2y_2)
\end{align*}
meaning if we exchange the values of $x_1$ and $x_2$ then these probabilities swap values.
The corresponding possibilities for the likelihood values are
\begin{align*}
    P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1+w)(1+y_1)(1+y_2); \\
    P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1-w)(1+y_1)(1+y_2); \\
    P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1-w)(1-y_1)(1-y_2); \\
    P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1+w)(1-y_1)(1-y_2);
\end{align*}
for site split $\{1\}$ and
\begin{align*}
        P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1+w)(1+y_1)(1+y_2); \\
    P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1-w)(1+y_1)(1+y_2); \\
    P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1-w)(1-y_1)(1-y_2); \\
    P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1+w)(1-y_1)(1-y_2);
\end{align*}
for site split $\{3\}$, which also both swap values when $x_1$ and $x_2$ are exchanged.

The same can be done for the splits $\{2\}$ and $\{1,2,3\}$ by exchanging $y_1$ and $y_2$ as well as $\{1,2\}$ and $\{1,3\}$ by exchanging both $x_1$ with $x_2$ and $y_1$ with $y_2$.
The split $\{1,3\}$ is unchanged by exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$.

Since exchanging $x_1$ and $x_2$ does not change the value of the log likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t}; \fullAncestralSplitPartitions(\tau_1,\tilde{t}))$, $x_1=x_2$ at the maximum.
The analogous statement holds for $y_1$ and $y_2$.
The Felsenstein topology does not admit this property, but, since we are interested in a lower bound for this topology, we simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.

%EM Here we are restating.
\topoInconsist*

Our approach seeks bounds for the terms inside the maximization in \eqref{eq:log_likelihood_simplified}.
We obtain
$$
\max_{t} \ \ell_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)) \le
    \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
    + \tilde{\ell}_{\tau^*,t^*}(\tau, \hat{t}; \fullAncestralSplitPartitions(\tau, \hat{t}))
$$
as an upper bound for the joint maximum of \eqref{eq:log_likelihood_simplified} using Gibbs's inequality
$$
\shannonDivergence_{\tau^*,t^*}(\tau,t) \le \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
$$
and
$$
\hat{t} = \argmax_{t} \ \tilde{\ell}_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)).
$$
Similarly, we have the lower bound
$$
\max_{t} \ \ell_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)) \ge
    \shannonDivergence_{\tau^*,t^*}(\tau,\hat{t})
    + \tilde{\ell}_{\tau^*,t^*}(\tau, \hat{t}; \fullAncestralSplitPartitions(\tau, \hat{t})).
$$

Assume $\tau^*=\tau_1$ and $t^*=\{x,y,y\}$.
For compactness, define
$$
p_{\emptyset} = P(\siteSplitRV=\emptyset \mid \tau=\tau_1,t=\{x,y,y\})
$$
with similar definitions for the remaining generating probabilities.
The full Farris tree log likelihood takes one of three values depending on branch lengths, which is due to site split $\{1,3\}$ (see Table~\ref{tab:likelihoods} for $\tau = \tau_1$).
We begin by writing out the middle case: for the ancestral state split $\{1\}$.
Directly substituting in \eqref{eq:site_pattern_likelihood}, the log likelihood is
\begin{align}
    \label{eq:farris_likelihood}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
    &=        p_{\emptyset}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \nonumber \\
    &\qquad + p_{1}          \cdot\log(1-x'^2+y'^2-x'^2y'^2) \nonumber \\
    &\qquad + p_{2}          \cdot\log(1+x'^2-y'^2-x'^2y'^2) \nonumber \\
    &\qquad + p_{3}          \cdot\log(1-x'^2+y'^2-x'^2y'^2) \nonumber \\
    &\qquad + p_{12}         \cdot\log(1-x'^2-y'^2+x'^2y'^2) \nonumber \\
    &\qquad + p_{13}         \cdot\log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \nonumber \\
    &\qquad + p_{23}         \cdot\log(1-x'^2-y'^2+x'^2y'^2) \nonumber \\
    &\qquad + p_{123}        \cdot\log(1+x'^2-y'^2-x'^2y'^2) \nonumber \\
    &\qquad - \log(8) \nonumber \\
    &\qquad + p_{\emptyset}  \cdot\log((1+x')^2   (1+w')(1+y')^2) \nonumber \\
    &\qquad + p_{1}          \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \nonumber \\
    &\qquad + p_{2}          \cdot\log((1+x')^2   (1+w')(1+y')(1-y')) \nonumber \\
    &\qquad + p_{3}          \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \nonumber \\
    &\qquad + p_{12}         \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y')) \nonumber \\
    &\qquad + p_{13}         \cdot\log((1+x')^2   (1-w')(1+y')^2) \nonumber \\
    &\qquad + p_{23}         \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y')) \nonumber \\
    &\qquad + p_{123}        \cdot\log((1+x')^2   (1+w')(1+y')(1-y')).
\end{align}
Seeking an upper bound, we use the facts that, for $x',y'\in[0,1]$,
\begin{align*}
1-x'^2+y'^2-x'^2y'^2 & = (1+x')(1-x')(1+y'^2) \le (1+x')(1-x')(1+y') \\
1+x'^2-y'^2-x'^2y'^2 & = (1+x'^2)(1+y')(1-y') \le (1+x')(1+y')(1-y') \\
1+x'^2+y'^2+x'^2y'^2 & = (1+x'^2)(1+y'^2) \le (1+x')(1+y') \\
4x'y' & = 2x' \cdot 2y' \le (1+x'^2)(1+y'^2) \le (1+x')(1+y').
\end{align*}
The term $p_{\emptyset}$ is bounded as
\begin{align*}
    p_{\emptyset} & = \log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
                  & \le \log(1+x'^2+y'^2+4x'y'+x'^2y'^2) \\
                  & \le \log(2(1+x')(1+y'))
\end{align*}
and $p_{13}$ is bounded as
\begin{align*}
    p_{13} & = \log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \\
                  & \le \log(1+x'^2+y'^2+x'^2y'^2) \\
                  & \le \log((1+x')(1+y')).
\end{align*}
Factoring and making these substitutions results in
\begin{align*}
&    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\})) \\
&\qquad \le      p_{\emptyset}  \cdot\log(2(1+x')(1+y'))
+ p_{1}          \cdot\log((1+x')(1-x')(1+y')) \\
    &\qquad\qquad + p_{2}          \cdot\log((1+x')(1+y')(1-y'))
+ p_{3}          \cdot\log((1+x')(1-x')(1+y')) \\
    &\qquad\qquad + p_{12}         \cdot\log((1+x')(1-x')(1+y')(1-y'))
+ p_{13}         \cdot\log((1+x')(1+y')) \\
    &\qquad\qquad + p_{23}         \cdot\log((1+x')(1-x')(1+y')(1-y'))
+ p_{123}        \cdot\log((1+x')(1+y')(1-y'))
- \log(8) \\
    &\qquad\qquad + p_{\emptyset}  \cdot\log((1+x')^2   (1+w')(1+y')^2)
+ p_{1}          \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \\
    &\qquad\qquad + p_{2}          \cdot\log((1+x')^2   (1+w')(1+y')(1-y'))
+ p_{3}          \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \\
    &\qquad\qquad + p_{12}         \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y'))
+ p_{13}         \cdot\log((1+x')^2   (1-w')(1+y')^2) \\
    &\qquad\qquad + p_{23}         \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y'))
+ p_{123}        \cdot\log((1+x')^2   (1+w')(1+y')(1-y')) \\
&\qquad =      p_{\emptyset}  \cdot\log(2)
+ \log(1+x')
+ (p_{1}+p_{3}+p_{12}+p_{23})\cdot\log(1-x') \\
&\qquad\qquad + \log(1+y')
+ (p_{2}+p_{12}+p_{23}+p_{123})\cdot\log(1-y') \\
&\qquad\qquad + (2-p_{1}-p_{3}-p_{12}-p_{23})\cdot\log(1+x')
+ (p_{1}+p_{3}+p_{12}+p_{23})\cdot\log(1-x') \\
&\qquad\qquad + (2-p_{2}-p_{12}-p_{23}-p_{123})\cdot\log(1+y')
+ (p_{2}+p_{12}+p_{23}+p_{123})\cdot\log(1-y') \\
&\qquad\qquad + (1-p_{13})\cdot\log(1+w')
+ p_{13}\cdot\log(1-w')
- \log(8) \\
&\qquad =      p_{\emptyset}  \cdot\log(2)
+ (3-p_{1}-p_{3}-p_{12}-p_{23})\cdot\log(1+x')
+ 2(p_{1}+p_{3}+p_{12}+p_{23})\cdot\log(1-x') \\
&\qquad\qquad + (3-p_{2}-p_{12}-p_{23}-p_{123})\cdot\log(1+y')
+ 2(p_{2}+p_{12}+p_{23}+p_{123})\cdot\log(1-y') \\
&\qquad\qquad + (1-p_{13})\cdot\log(1+w')
+ p_{13}\cdot\log(1-w')
- \log(8).
\end{align*}
To make matters further compact, let
\begin{equation}
    \begin{aligned}
        a_{1} &= 3-p_{1}-p_{3}-p_{12}-p_{23}, \\
        a_{2} &= 2(p_{1}+p_{3}+p_{12}+p_{23}), \\
        a_{3} &= 3-p_{2}-p_{12}-p_{23}-p_{123}, \\
        a_{4} &= 2(p_{2}+p_{12}+p_{23}+p_{123}), \\
    \end{aligned}
    \label{eq:a_const}
\end{equation}
where
\begin{equation*}
\begin{split}
&    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\})) \\
